### 2022-03-29 paper/localpsf.tex
editing intro

@@

@@

The basic idea behind our method is impulse response interpolation. The impulse response, $\phi_x$, associated with a point $x$ is the function that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.5]{bct1.eps}
	\end{center}
	\caption{Hierarchical matrix block structure for the Stokes Hessian. Red blocks are dense. All other blocks are low rank.
	}
	\label{fig:hmatrix_intro}
\end{figure}

@@

In this paper, we build on previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format. Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to make the following improvements:
\begin{itemize}
	\item PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while the method we propose is based on an assumption that we call ``normalized local mean displacement invariance'' (see Figure \ref{fig:mean_displacement_invariance}). This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). 
	\item PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. 
\end{itemize}
In exchange for these improvements, the proposed approximation cannot be written as a weighted sum of convolution operators, hence we cannot apply this approximation to vectors via the FFT. However, since our approximation is an H-matrix, it can be applied to vectors efficiently via H-matrix methods.

Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}, which we do not use. These methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. One issue with this approach is that, while these methods are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more efficient in the future.

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. 
%We write $L^2(\Omega)'$ to denote the topological dual of $L^2(\Omega)$. 
%The linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. 
The action of $\Aop$ may be extended to distributions if the kernel $\Phi$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy,
	\end{align}
\end{subequations}
%\begin{align*}
%\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx dy \\
%&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx \\
%&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy 
%%= \left(v,\genericdistribution\left(\Phi(y,\cdot)\right)\right)_{L^2(\Omega)},
%\end{align*}
where $\Phi(y,\cdot)$ denotes the function $x \mapsto \Phi(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Phi$ is sufficiently regular such that the function $y \mapsto \rho\left(\Phi(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the last line in the derivation is well-defined. Hence, we use the last line of this derivation to \emph{define} the action of $\Aop$ on the distribution $\rho$ as follows:
\begin{equation}
\label{eq:action_on_distribution}
\left(\Aop \genericdistribution^*\right)(v) := \int_\Omega v(y) \genericdistribution\left(\Phi(y, \cdot)\right) dy.
\end{equation}
Since $\genericdistribution^*$ may not exist, here ``$\Aop \genericdistribution^*$'' is simply a notation that means ``the action of $\Aop$ on the distribution $\rho$,'' per \eqref{eq:action_on_distribution}.

>>>>

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.4]{bct1.eps}
	\end{center}
	\caption{Hierarchical matrix block structure for the Stokes Hessian. The red blocks (tiny) are dense. All other blocks are low rank.
	}
	\label{fig:hmatrix_intro}
\end{figure}

@@

Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}, which we do not use. These alternative methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. One issue with this approach is that, while these methods are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more efficient in the future.

@@

The basic idea behind our method is impulse response interpolation. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

@@

In this paper, we build on previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format. Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to make the following improvements:
\begin{itemize}
	\item PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while the method we propose is based on an assumption that we call ``local mean displacement invariance'' (see Figure \ref{fig:mean_displacement_invariance}). This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). 
	\item PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. 
\end{itemize}
In exchange for these improvements, the proposed approximation cannot be written as a weighted sum of convolution operators, hence we cannot apply this approximation to vectors via the FFT. However, since our approximation is an H-matrix, it can be applied to vectors efficiently via H-matrix methods.

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. 
%We write $L^2(\Omega)'$ to denote the topological dual of $L^2(\Omega)$. 
%The linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$.

The action of $\Aop$ may be extended to distributions if the kernel $\Phi$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy,
	\end{align}
\end{subequations}
%\begin{align*}
%\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx dy \\
%&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx \\
%&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy 
%%= \left(v,\genericdistribution\left(\Phi(y,\cdot)\right)\right)_{L^2(\Omega)},
%\end{align*}
where $\Phi(y,\cdot)$ denotes the function $x \mapsto \Phi(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Phi$ is sufficiently regular such that the function $y \mapsto \rho\left(\Phi(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the last line in the derivation is well-defined. Hence, we use the last line of this derivation to \emph{define} the action of $\Aop$ on the distribution $\rho$ as follows:
\begin{equation}
\label{eq:action_on_distribution}
\left(\Aop \genericdistribution^*\right)(v) := \int_\Omega v(y) \genericdistribution\left(\Phi(y, \cdot)\right) dy.
\end{equation}
Since $\genericdistribution^*$ may not exist, here ``$\Aop \genericdistribution^*$'' is simply a notation that means ``the action of $\Aop$ on the distribution $\rho$,'' per \eqref{eq:action_on_distribution}.
